Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-19 MS_thesis.tex
Rephrase abstract

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit and can help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for evolutionary agent-based models and empirical studies.
}

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in a world where population sizes  routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit and can help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for evolutionary agent-based models and empirical studies.
}
